{memes}: an R interface to the MEME Suite

### 2020-09-22 manuscript.Rmd
Almost final draft

---
title: '{memes}: an R interface to the MEME Suite'
author:
- ORCID: null
  name: Spencer L. Nystrom
- name: Daniel J. McKay
output: word_document
---

```{r setup}
library(dremeR)
library(magrittr)
```

@@

Many of the MEME Suite tools, such as MEME, DREME, or TomTom return data which
describe properties of motifs. For example, MEME and DREME return de-novo
discovered motifs from query sequences along with statistical information about
their enrichment. The `universalmotif` data structure is a powerful
R/Bioconductor representation for motif matrices and their associated metadata \cite{universalmotif}. 
However, `universalmotif` objects are not ideal for performing bulk manipulation
or queries of metadata, nor do they support complex data structures such as
nested elements. Many MEME Suite analysis pipelines involve identifying motifs
which are further analyzed by multiple tools, creating additional metadata which
cannot be easily tracked in bulk using `universalmotif` format. Thus
an ideal representation for these data is one which can hold an unlimited amount
of metadata, contain complex heirarchies, and be easily manipulated using
standard analysis tools. `memes` therefore extends the `universalmotif` format
by providing data.frames holding motif metadata which retain links to their
`universalmotif` representation. These universalmotif
data.frames combine the advantages of the `universalmotif` framework for motif
manipulation with the extensibility and flexibility of R data.frames. These
`universalmotif` data.frames form the basis for a majority of `memes` data
outputs. These structures are also valid input types to `memes` functions, and
when used as such, output data are appended as new columns to the input data,
ensuring data provenance.

@@

During analysis of multiomic data, users often uncover different functional
categories of genomic regions through integration of multiple datasets. In these
instances, motif analysis acts as a powerful tool to examine whether differences
in motif content is associated with membership of distinct categories. In some
cases, performing differential motif analysis (using one category as the "test"
set and another as the "control") can uncover biologically relevant motifs which
are enriched in one category relative to another. Although the MEME Suite allows
differential enrichment testing, it does not inheritly provide a mechanism for
analyzing groups of sequences in parallel, or performing motif analysis with any
understanding of data categories. The `memes` framework enables "data-aware"
motif analysis workflows by allowing named lists of `Biostrings` objects as
input to each function. If using `GenomicRanges`, users can split peak data on a
metadata column using the base R `split` function, and use this result as input
to `get_sequence`, which will produce a list of `BioStrings` objects where each
entry is named after each data category from the split column. When a list is
used as input to a `memes` function, it runs the corresponding MEME Suite tool
for each object in the list. Users can also pass the name(s) of a category to
the `control` argument to automatically enable differential analysis of all
other list members against the control category sequences.  Therefore, `memes`
enables complex data-aware differential motif analysis workflows using simple
syntax to extend the capabilities of the MEME Suite.

@@

## Containerized analysis maximizes availability and facilitates reproducibility
`memes` relies on a locally installed version of the MEME Suite which is
accessed by the users R process. Although R is available on Windows, Mac, and
Linux operating systems, the MEME Suite is incompatible with Windows, limiting
its adoption by Windows users. Additional barriers also exist to installing a
local copy of the MEME Suite on compatible systems. The MEME Suite relies on
several system-level dependencies whose installation can be difficult for novice
users. Finally, some tools in the MEME Suite use python to generate shuffled
control sequences for analysis, which presents a reproducibility issue as the
random number generation algorithm changed from python2.7 to python3. The MEME
Suite will build and install on both python2.7 and python3 systems, therefore
without careful consideration of this behavior, the same code run on two systems
may not produce identical results, even if using the same major version of the
MEME Suite. In order to democratize access to the MEME Suite on unsupported
operating systems, and to facilitate reproducibile motif analysis using `memes`, 
we have also developed docker and singularity (**NOTE:** singularity is broken
for `rocker/rstudio`, so maybe omit this until fixed) containers containing a
preinstalled version of the MEME Suite along with an R/Bioconductor analysis
environment including the most recent version of `memes` and its dependencies.

# Future developments:
 - Additional meme suite tools
 - more data visualizations
 - better support for PSP generation
 - Better support for amino-acid sequence usage (please contact us with feature requests)

# Acknowledgements
 - Sean Davis automated docker builds
 - Megan Justice, feedback, feature requests

>>>>

---
title: '{memes}: an R interface to the MEME Suite'
author:
- ORCID: null
  name: Spencer L. Nystrom
- name: Daniel J. McKay
output: 
 - word_document:
   fig.caption: yes
---

```{r setup, echo = FALSE}
# Figure caption: https://stackoverflow.com/questions/31926623/figures-captions-and-labels-in-knitr
suppressPackageStartupMessages(library(memes))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(GenomicRanges))
suppressPackageStartupMessages(library(ggplot2))
drake::r_make()
```

@@

Many of the MEME Suite tools, such as MEME, DREME, or TomTom return data which
describe properties of motifs. For example, MEME and DREME return de-novo
discovered motifs from query sequences along with statistical information about
their enrichment. The `universalmotif` data structure is a powerful
R/Bioconductor representation for motif matrices and their associated metadata \cite{universalmotif}. 
However, `universalmotif` objects are not ideal for performing bulk manipulation
or queries of metadata, nor do they support complex data structures such as
nested elements. Many MEME Suite analysis pipelines involve identifying motifs
which are further analyzed by multiple tools, creating additional metadata which
cannot be easily tracked in bulk using `universalmotif` format. Thus
an ideal representation for these data is one which can hold an unlimited amount
of metadata, contain complex hierarchies, and be easily manipulated using
standard analysis tools. `memes` therefore extends the `universalmotif` format
by providing data.frames holding motif metadata which retain links to their
`universalmotif` representation. These universalmotif
data.frames combine the advantages of the `universalmotif` framework for motif
manipulation with the extensibility and flexibility of R data.frames. These
`universalmotif` data.frames form the basis for a majority of `memes` data
outputs. These structures are also valid input types to `memes` functions, and
when used as such, output data are appended as new columns to the input data,
ensuring data provenance.

@@

During analysis of multiomic data, users often uncover different functional
categories of genomic regions through integration of multiple datasets. In these
instances, motif analysis acts as a powerful tool to examine whether differences
in motif content is associated with membership of distinct categories. In some
cases, performing differential motif analysis (using one category as the "test"
set and another as the "control") can uncover biologically relevant motifs which
are enriched in one category relative to another. Although the MEME Suite allows
differential enrichment testing, it does not inherently provide a mechanism for
analyzing groups of sequences in parallel, or performing motif analysis with any
understanding of data categories. The `memes` framework enables "data-aware"
motif analysis workflows by allowing named lists of `Biostrings` objects as
input to each function. If using `GenomicRanges`, users can split peak data on a
metadata column using the base R `split` function, and use this result as input
to `get_sequence`, which will produce a list of `BioStrings` objects where each
entry is named after each data category from the split column. When a list is
used as input to a `memes` function, it runs the corresponding MEME Suite tool
for each object in the list. Users can also pass the name(s) of a category to
the `control` argument to automatically enable differential analysis of all
other list members against the control category sequences.  Therefore, `memes`
enables complex data-aware differential motif analysis workflows using simple
syntax to extend the capabilities of the MEME Suite.

@@

## Containerized analysis maximizes availability and facilitates reproducibility
`memes` relies on a locally installed version of the MEME Suite which is
accessed by the users R process. Although R is available on Windows, Mac, and
Linux operating systems, the MEME Suite is incompatible with Windows, limiting
its adoption by Windows users. Additional barriers also exist to installing a
local copy of the MEME Suite on compatible systems. The MEME Suite relies on
several system-level dependencies whose installation can be difficult for novice
users. Finally, some tools in the MEME Suite use python to generate shuffled
control sequences for analysis, which presents a reproducibility issue as the
random number generation algorithm changed from python2.7 to python3. The MEME
Suite will build and install on both python2.7 and python3 systems, therefore
without careful consideration of this behavior, the same code run on two systems
may not produce identical results, even if using the same major version of the
MEME Suite. In order to democratize access to the MEME Suite on unsupported
operating systems, and to facilitate reproducible motif analysis using `memes`, 
we have also developed docker and singularity (**NOTE:** singularity is broken
for `rocker/rstudio`, so maybe omit this until fixed) containers containing a
preinstalled version of the MEME Suite along with an R/Bioconductor analysis
environment including the most recent version of `memes` and its dependencies.
As new container versions are released, they are version tagged and stored to
ensure reproducibility while allowing updates to the container.

# Results

Here we briefly highlight each of `memes` current features for analyzing
ChIP-seq data. Additional detailed walkthroughs for each supported MEME Suite
tool, and a complete worked example using `memes` to analyze ChIP-seq data can be found
in the `memes` vignette and the package website
([snystrom.github.io/memes](snystrom.github.io/memes)). In the following
example, we reanalyze recent work examining the causal relationship between the
binding of the transcription factor E93 to changes in chromatin accessibility
during *Drosophila* wing development \cite{Nystrom, Niederhuber 2020}. Here, we
utilize ChIP-seq peak calls for E93 which have been annotated according to the
change in chromatin accessibility observed before & after E93 binding. These
data are an idiomatic example of range-based genomic data (E93 ChIP peaks)
containing additional groupings (the chromatin accessibility response following
DNA binding) whose membership may be influenced by differential motif content.
We show how `memes` syntax allows complex analysis designs, how `memes`
utilities enable deep interrogation of results, and how `memes` flexible data
structures empower users to integrate the `memes` workflow with tools offered by
other R/Bioconductor packages.

```{r, echo = F}
drake::loadd(chip_results)
```

The aforementioned ChIP-seq peaks are stored as a GRanges object with a metadata
column (`e93_chromatin_response`) indicating whether chromatin accessibility
tends to increase ("Increasing"), decrease ("Decreasing"), or remain unchanged
("Static") following E93 binding.
```{r}
head(chip_results, 3)
```

Using the `get_sequence` function, GRanges objects are converted into
DNAStringSet outputs.
```{r}
dm.genome <- BSgenome.Dmelanogaster.UCSC.dm3::BSgenome.Dmelanogaster.UCSC.dm3

all_sequences <- chip_results %>% 
  get_sequence(dm.genome)
```

```{r}
head(all_sequences)
```
In order to perform analysis within different groups of peaks, the GRanges object can be `split` on a metadata column before input to `get_sequence`.
```{r}
sequences_by_response <- chip_results %>% 
  split(mcols(.)$e93_chromatin_response) %>% 
  get_sequence(dm.genome)

```

This produces in a BStringSetList where list members contain a DNAStringSet for each group of sequences.
```{r}
head(sequences_by_response)
```

## Denovo motif analysis
The DREME tool can be used to discover short, novel motifs present in a set of
input sequences relative to a control set. The `runDreme` command is the `memes`
interface to the DREME tool. `runDreme` syntax enables users to produce complex
discriminative analysis designs using intuitive syntax. Examples of possible
designs and their syntax are compared below.

```{r, eval=F}
# use all sequences vs shuffled background
runDreme(all_sequences, control = "shuffle")

# for each response category, discover motifs against a shuffled background set
# Produces:
#  - Increasing vs Shuffled Increasing
#  - Decreasing vs Shuffled Decreasing
#  - Static vs Shuffled Static
runDreme(sequences_by_response, control = "shuffle")

# Search for motifs enriched in the "Increasing" peaks relative to "Decreasing" peaks
# Produces:
#  - Increasing vs Decreasing
runDreme(sequences_by_response$Increasing, control = sequences_by_response$Decreasing)

# Use the "Static" response category as the control set to discover motifs 
# enriched in each remaining category relative to static sites,
# Produces:
#  - Increasing vs Static
#  - Decreasing vs Static
runDreme(sequences_by_response, control = "Static")

# Combine the "Static" and "Increasing" sequences and use as a background set to
# discover motifs enriched in the remaining categories relative
# Produces:
#  - Decreasing vs Static+Increasing
runDreme(sequences_by_response, control = c("Static", "Increasing"))
```

In order to discover motifs associated with dynamic changes to chromatin
accessibility, we use static sites as the control set to discover motifs
enriched in both increasing and decreasing sites.

```{r, eval=F}
dreme_vs_static <- runDreme(sequences_by_response, control = "Static")
```

```{r, echo=FALSE}
# I secretly cache this to decrease build-time of the manuscript
drake::loadd(dreme_vs_static)
```

```{r, fig.height=4, fig.width=7, echo = FALSE}
dreme_fig_top <- purrr::map(dreme_vs_static, "motif") %>% 
  purrr::map(universalmotif::view_motifs) %>% 
  purrr::imap(~{
    .x +
      ggtitle(paste(.y, "Motifs")) +
      theme(plot.title = element_text(hjust = 0.5, size = 14))
  }) %>% 
  patchwork::wrap_plots(ncol = 2) 
```
```{r, echo=FALSE}

# Require that the minimum overlap for similarity metric is equal to the
# smallest possible overlap if the shortest motif was completely contained
# within the largest motif.
get_min_overlap <- function(df){
  min(df$length) / max(df$length)
}

get_motif_correlation <- function(df, method = "PCC"){
  df$motif %>% 
    universalmotif::compare_motifs(min.overlap = get_min_overlap(df), method = method)
}

plot_correlation_heatmap <- function(mat, name){

  min_color <- round(min(mat) - 0.1, digits = 1)
  colors <- circlize::colorRamp2(c(min_color, 1), c("white", "dodgerblue3"))

  rownames(mat) <- NULL
  ComplexHeatmap::Heatmap(mat,
                          col = colors, 
                          name = "Pearson Correlation", 
                          column_names_rot = 45,
                          column_title = paste(name, "Motif Similarity")
                          )

}
```

```{r, fig.height=5, fig.width=10, echo = FALSE}
dreme_fig_bottom <- dreme_vs_static %>% 
  purrr::map(get_motif_correlation) %>% 
  purrr::imap(plot_correlation_heatmap) %>% 
  # capture heatmaps as grobs
  purrr::map(~{grid::grid.grabExpr(draw(.x))}) %>% 
  patchwork::wrap_plots(nrow = 1) 
```

These data are readily visualized using the `universalmotif::view_motifs`
function. Visualization of the results reveals two distinct motifs associated
with decreasing and increasing sites \ref{fig:dec_v_inc_dreme}. In this
analysis, the *de-novo* discovered motifs within each category appear visually
similar to each other, however, the MEME Suite does not provide a mechanism to
compare groups of motifs based on all pairwise similarity metrics. We utilized
the `universalmotif::compare_motifs` function to compute Pearson correlation
coefficients for each set of motifs to quantitatively assess motif similarity
\ref{fig:dec_v_inc_dreme}.

```{r dec_v_inc_dreme, fig.height=10, fig.width=10, echo=F, fig.cap="\\label{fig:dec_v_inc_dreme}PWMs of discovered motifs in Decreasing (A) and Increasing (B) sites. C,D Pearson correlation heatmaps comparing similarity of Decreasing (C) and Increasing (D) motifs."}
patchwork::wrap_plots(dreme_fig_top, dreme_fig_bottom, ncol = 1) +
  patchwork::plot_annotation(tag_levels = "A")
```
Transcription factors often bind with sequence specificity to regulate activity
of nearby genomic elements. Therefore, comparison of *de-novo* motifs with known
DNA binding motifs for transcription factors can be a key step in identifying
regulators of genomic elements. The TomTom tool is used to compare unknown
motifs against a list of known motifs to identify matches, `memes` provides the
`runTomTom` function as an interface to this tool. By passing the results from
`runDreme` into `runTomTom` and searching within a database of *Drosophila* transcription
factor motifs, we can identify candidate transcription factors which may bind
the motifs associated with increasing and decreasing chromatin accessibility.

```{r}
dreme_vs_static_tomtom <- runTomTom(dreme_vs_static, database = "data/flyFactorSurveyCleaned.meme", dist = "ed")
```
Using this approach, the results can be visualized using 
the `view_tomtom_hits()` function to visually inspect the matches
assigned by TomTom, providing a simple way to assess the quality of the match. 
A representative plot of these data is shown in \ref{fig:tomtom_example}, while
the full results can be viewed in \ref{tab:tomtom_table}. We
uncover that the motifs associated with decreasing sites are highly similar to
the E93 motif. Conversely, motifs found in increasing sites match the
transcription factors br, l(3)neo38, gl, and lola \ref{tab:tomtom_table}. These
data support the hypothesis that E93 binding to its motif may play a larger role
in chromatin closing activity, while binding of the transcription factor br is
associated with chromatin opening.
```{r tomtom_example, fig.height=3, fig.width=7.5, fig.cap="\\label{fig:tomtom_example}Representative plots generated by view_tomtom_hits showing the top hit for motifs discovered in decreasing (A) and increasing (B) peaks."}
plot_top_hit <- function(x){
  x %>% 
    view_tomtom_hits(1)
}

top_hit_grid <- function(x, title){
  x %>% 
    patchwork::wrap_plots(nrow = 1)
}

# Plotting every hit:
# decided to just use a representative sample insted
#dreme_vs_static_tomtom %>% 
#  purrr::imap(~{plot_top_hit(.x) %>% top_hit_grid(., .y)}) %>% 
#  patchwork::wrap_plots(ncol = 1)

dreme_vs_static_tomtom %>% 
  purrr::imap(~{
    .x %>% 
      dplyr::mutate(name = paste0(.y, " ", name)) %>% 
      update_motifs() %>% 
      plot_top_hit() %>% 
      .[[2]]
    }) %>% 
  patchwork::wrap_plots() +
    patchwork::plot_annotation(tag_levels = "A")
```
```{r}
library(gt)

dreme_vs_static_tomtom %>% 
  dplyr::bind_rows(.id = "category") %>% 
  dplyr::mutate(category = paste(category, "Motifs")) %>% 
  dplyr::arrange(category, rank) %>% 
  dplyr::select(category, name, best_match_altname) %>% 
  dplyr::rename("Best Match TF" = best_match_altname) %>% 
  gt(rowname_col = "name",
     groupname_col = "category",
     lab = "tab:tomtom_table")

```

## Motif enrichment testing using runAme

Discovery and matching of *de-novo* motifs is only one way to find candidate
transcription factors within target sites. Indeed, in many instances, requiring
that a motif is recovered *de-novo* is not ideal, as these approaches are
"discovery based" as opposed to hypothesis driven. Another approach, implemented
by the AME tool, is to search for known motif instances in target sequences and
test for their overrepresentation relative to a background set of sequences
\cite{ame}. The `runAme()` function is the `memes` interface to the AME tool. It
accepts a set of sequences as input an control sets, and will peform enrichment
testing using a provided motif database for occurrances of each provided motif.

A major limitation of this approach is that transcription factors containing
similar families of DNA binding domain often possess highly similar motifs,
making it difficult to identify the "true" binding factor associated with an
overrepresented motif. Additionally, when searching for matches against a motif
database, AME must account for multiple testing, therefore using a larger than
necessary motif database can produce a large multiple testing penalty, limiting
sensitivity of detection. One way to overcome these limitations is to limit the
transcription factor motif database to include only motifs for transcription
factors expressed in the sample of interest. Accounting for transcription factor
expression during motif analysis has been demonstrated to increase the
probability of identifying biologically relevant transcription factor candidates
\cite{hector franco's STAR paper, others?}.

The `memes` universalmotif data.frame structure can be used to integrate
expression data with a motif database to remove entries for transcription
factors which are not expressed. To do so, we import a *Drosophila*
transcription factor motif database generated by the Fly Factor Survey and
convert to universalmotif data.frame format \cite{fly factor}. In this database,
the `altname` column stores the gene symbol.

```{r}
fly_factor <- universalmotif::read_meme("data/flyFactorSurveyCleaned.meme") %>% 
  as_universalmotif_dataframe()
```
```{r, echo = FALSE}
head(fly_factor, 3)
```
Next, we import a pre-filtered list of genes expressed in a timecourse of
**Drosophila** wing development.
```{r}
wing_expressed_genes <- read.csv("data/wing_expressed_genes.csv")
```

```{r, echo = FALSE}
n_fly_factor_expressed <- fly_factor %>% 
  dplyr::filter((altname %in% wing_expressed_genes$symbol)) %>% 
  nrow()
n_fly_factor_unexpressed <- fly_factor %>% 
  dplyr::filter(!(altname %in% wing_expressed_genes$symbol)) %>% 
  nrow()
```

Finally, we subset the motif database to only expressed genes using `dplyr`
data.frame subsetting syntax (note that base R subsetting functions operate
equally well on these data structures), then convert the data.frame back into
`universalmotif` format using `as_universalmotif`. This filtering step removes
`r (n_fly_factor_unexpressed / nrow(fly_factor)) %>% {. * 100} %>% round()`% of
entries from the original database, greatly reducing the multiple-testing
correction.

```{r}
fly_factor_expressed <- fly_factor %>% 
  dplyr::filter(altname %in% wing_expressed_genes$symbol) %>% 
  as_universalmotif()
```

`runAme` syntax is identical to `runDreme` in that discriminative designs can be
constructed by calling list entries by name. Data can be visualized using
`ame_plot_heatmap()`, revealing that the E93 motif is strongly enriched at
Decreasing sites, while l(3)neo38, lola, and br are enriched in Increasing
sites, supporting the *de-novo* discovery results \ref{fig:ame_heatmap}.
Additionally, AME detects several other transcription factor motifs which
distinguish decreasing and increasing sites, providing additional clues to
potential factors which also bind E93 ChIP peaks to affect chromatin
accessibility.
```{r}
ame_vs_static <- runAme(sequences_by_response, control = "Static", database = fly_factor_expressed)
```

```{r ame_heatmap, fig.height=3, fig.width=7.5, echo = FALSE, fig.cap="\\label{fig:ame_heatmap} A heatmap of the -log10(adj.pvalues) from AME enrichment testing within increasing and decreasing sites."}
ame_vs_static %>% 
  dplyr::bind_rows(.id = "category") %>% 
  dplyr::mutate(category = paste0(category, "\nAccessibility")) %>% 
  dplyr::group_by(category, motif_alt_id) %>% 
  dplyr::filter(adj.pvalue == min(adj.pvalue)) %>% 
  ame_plot_heatmap(group = category, id = motif_alt_id) +
    labs(x = "Transcription Factor",
         y = "ChIP Peak Type")
```
### Motif matching using runFimo

A striking result from these analysis is that the E93 motif is so strongly
enriched within E93 ChIP peaks which decrease in accessibility. Significance of
motif enrichment can be driven by several factors, such as quality of the query
motif relative to the canonical motif or differences in motif number in one
group relative to other groups. These questions can be explored directly by
identifying motif occurrances in target regions and examining their properties.
FIMO allows users to match motifs in input sequences while returning information
about the quality of each match in the form of a quantitative score \cite{fimo}.

In order to examine the properties of the E93 motif between different ChIP
peaks, we scan all E93 ChIP peaks for matches to the Fly Factor Survey E93 motif
using `runFimo`. Results are returned as a GRanges object containing the
positions of each motif. Setting `skip_matched_sequence` to `FALSE` adds the
sequence of the detected motif to the output.

```{r}
e93_fimo <- runFimo(all_sequences, fly_factor_expressed$Eip93F_SANGER_10, skip_matched_sequence = FALSE, thresh = 1e-3)
```

Using `plyranges`, matched motifs can be joined with the metadata of the ChIP peaks with which they intersect.
```{r}
e93_fimo %<>% 
  plyranges::join_overlap_intersect(chip_results)
```

```{r, echo=F}
e93_stack <- e93_fimo %>% 
  data.frame %>%  
  split(.$e93_chromatin_response) %>% 
  purrr::imap(~{
    universalmotif::create_motif(.x$matched_sequence, name = paste(.y, "E93"))
  }) %>% 
  rev %>% 
  {
    m <- .
    e93_flyfactor <- fly_factor_expressed$Eip93F_SANGER_10
    e93_flyfactor@name <- "E93 FlyFactorSurvey"

    m$e93 <- e93_flyfactor
    return(m)
  } %>% 
  rev %>% 
  universalmotif::view_motifs()
```
```{r, echo=F}
n_motifs <- chip_results %>% 
  plyranges::mutate(has_match = ifelse(. %over% e93_fimo, "Match", "No Match")) %>% 
  data.frame %>% 
  ggplot(aes(e93_chromatin_response)) +
    geom_bar(aes(fill = forcats::fct_rev(has_match)), position = "fill") +
     scale_fill_manual(values = c("Match" = "firebrick",
                                  "No Match" = "Black")) +
     labs(fill = "E93 Motif Match",
          y = "Fraction of Sites",
          x = "Response to E93 binding") 
```
```{r, fig.height=5, fig.width=3, echo=F}
motif_quality <- e93_fimo %>% 
  data.frame %>% 
  ggplot(aes(e93_chromatin_response, score)) +
    geom_boxplot(aes(fill = e93_chromatin_response), notch = TRUE) +
    theme_bw() +
    labs(y = "FIMO Score",
         x = "Chromatin Response in E93 ChIP Peak") +
    theme(legend.position = "none")
```

Using this approach we can deeply examine the properties of the E93 motif within
each chromatin response category. First, by counting the number of E93 motifs
within each category, we demonstrate that Decreasing sites are more likely than
increasing or static sites to contain an E93 motif \ref{fig:fimo}A. We extend
these observations by rederiving position-weight matrices from sequences
matching the E93 motif within each category, allowing visual inspection of motif
quality across groups \ref{fig:fimo}B. Notably, differences in quality at base
positions 8-12 appear to distinguish increasing from decreasing motifs, where
decreasing motifs are more likely to have strong A bases at positions 8 and 9,
while E93 motifs from increasing sites are more likely to have a T base pair at
position 12 \ref{fig:fimo}B. Examination of bulk FIMO scores (where higher
scores represent motifs more similar to the reference) also reveals differences
in E93 motif quality between groups, in particular, E93 motifs from Decreasing
sites have higher scores \ref{fig:fimo}C. Together, these data demonstrate that
a key distinguishing factor between whether a sites will decrease or increase in
chromatin accessibility following E93 binding is the number and quality of E93
motifs at that site.

In summary, `memes` establishes a powerful motif analysis environment by
leveraging the speed and utility of the MEME Suite set of tools in conjunction
with the flexible and extensive R/Bioconductor package landscape.
```{r fimo, fig.width=7.5, fig.height=9, fig.cap="\\label{fig:fimo}A. Stacked barplot showing fraction of each chromatin response category containing E93 motif matches. B. PWMs generated from E93 motif sequences detected in each chromatin response category. C. Boxplot of FIMO Score for each E93 motif within each chromatin response category. Outliers are plotted as distinct points."}
#patchwork::wrap_plots(n_motifs, motif_quality, patchwork::guide_area(), e93_stack, widths = c(0.2, 0.3, 0.5))
patchwork::wrap_plots(n_motifs, patchwork::guide_area(), 
                      e93_stack, motif_quality, 
                      ncol = 2, widths = c(0.6, 0.4),
                      guides = 'collect') +
    patchwork::plot_annotation(tag_levels = "A")
```

# Availability and Future developments:

The `memes` package source code and installation instructions are available on
github: [github.com/snystrom/memes](github.com/snystrom/memes). Documentation is
stored in the package vignettes, and also available at the package website:
[snystrom.github.io/memes](snystrom.github.io/memes). The `memes_docker`
containers are available from DockerHub
[https://hub.docker.com/r/snystrom/memes_docker](https://hub.docker.com/r/snystrom/memes_docker),
and their source code is hosted at github:
[https://github.com/snystrom/memes_docker](https://github.com/snystrom/memes_docker).

This manuscript was automatically generated using Rmarkdown within the
`memes_docker` container (TODO: ADD VERSION). Its source code, raw data, and
instructions to reproduce all analysis can be found at
[github.com/snystrom/memes_paper/](github.com/snystrom/memes_paper/).

In the future we hope to add additional data visualizations for examining motif
positioning within features. We will continue to add support for additional MEME Suite
tools in future versions of the package. Finally, we hope to improve the `memes`
tooling for analyzing amino-acid motifs, which although fully supported by our
current framework, may require extra tools which we have not considered.

# Acknowledgements
We would like to acknowledge Megan Justice, and members of the McKay lab for
helpful comments and feedback.
